---
title: "ST405 Mini Project"
author: "S/18/841"
date: "2024-04-05"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```






**Introduction**

The National Health and Nutrition Examination Survey (NHANES) conducted by the Centers for Disease Control and Prevention (CDC) gathers extensive health and nutritional information from a diverse U.S. population. In this study, we narrow our focus to predicting respondents' age by extracting a subset of features from the NHANES dataset. These features include physiological measurements, lifestyle choices, and biochemical markers, hypothesized to correlate strongly with age.

**Methodology**

**Dataset Description:**
The dataset used in this study is derived from NHANES 2013-2014, comprising various health and nutritional data from the U.S. population. Key variables include "SEQN" (Respondent Sequence Number), "age_group" (Respondent's Age Group), "RIDAGEYR" (Respondent's Age), "RIAGENDR" (Respondent's Gender), "PAQ605" (Engagement in Physical Activities), "BMXBMI" (Body Mass Index), "LBXGLU" (Blood Glucose after Fasting), "DIQ010" (Diabetes Status), "LBXGLT" (Oral Glucose Tolerance Test), and "LBXIN" (Blood Insulin Levels).

**Statistical Methods Employed:**
This analysis employs Exploratory Factor Analysis (EFA) and Confirmatory Factor Analysis (CFA). EFA uncovers underlying patterns and correlations among variables, while CFA validates the factor structure and assesses model fit. Descriptive statistics such as means, standard deviations, and correlations are computed to summarize the dataset and examine relationships between variables.

```{r}
#Libraries
library(tidyverse)
library(ggplot2)
library(psych)
library(corrplot)
library(ggcorrplot)
library(GPArotation)
library(nFactors)
library(factoextra)
library(psych)
library(lavaan)
```


```{r}
###Data Loading and Inspection
nhanes_data <- read.csv("../Data/NHANES_age_prediction.csv")
```


```{r}
# View the structure of the dataset
str(nhanes_data)
```


```{r}
# View summary statistics of the dataset
summary(nhanes_data)
```


```{r}
# View the first few rows of the dataset
head(nhanes_data)
```

```{r}
#Dimensions of the dataset
dim(nhanes_data)
```


```{r}
# Check for missing values
colSums(is.na(nhanes_data))

```

```{r}
###Data Preprocessing
# Select only numeric variables from the NHANES dataset
numerical_nhanes_data <- nhanes_data[, sapply(nhanes_data, is.numeric)]

# Scale the numeric variables
normalized_data <- scale(numerical_nhanes_data)

# View the first few rows of the normalized data
head(normalized_data)

```
```{r}
#Compute the correlation matrix
cor_matrix<-cor(normalized_data)
```

```{r}
#Visualize the correlation matrix
ggcorrplot(cor_matrix)
```

```{r}

```

```{r}
# Compute eigenvalues
eigen_values<- eigen(cor_matrix)
eigen_values
```
```{r}
# Principal Component Analysis (PCA)
PCA <- princomp(cor_matrix)
PCA
```
```{r}
summary(PCA)
```
```{r}
# Visualize the eigenvalues
fviz_eig(PCA,addlabels=TRUE)
```
```{r}
# Perform parallel analysis for factor extraction
fa.parallel(normalized_data, fm = "pa", fa = "fa")
```
```{r}
# Compute covariance matrix
covariance_matrix <- cov(normalized_data)
covariance_matrix
```
```{r}
# Kaiser-Meyer-Olkin (KMO) Test
KMO(r=nhanes_data[,-c(2)])

cor_matrix <- cor_matrix[,KMO(cor_matrix)$MSAi>0.5]
round(KMO(cor_matrix)$MSA,2)
```
```{r}
# Perform Bartlett's test of sphericity
cortest.bartlett(normalized_data)
```
```{r}
# Compute the determinant of the covariance matrix
det(covariance_matrix)
```
```{r}
# Compute proportion of variance explained by each principal component
Pop_var_exp <- eigen_values$values/sum(eigen_values$values)*100
round(Pop_var_exp,3)
```
```{r}
# Total variance explained by all principal components
sum(eigen_values$values)
```
```{r}
# Sum of proportion of variance explained
sum(Pop_var_exp)
```
```{r}
# Factor Loadings using PCA
PCA$loadings[,1:4]
```

```{r}
# Visualize the variables in PCA
fviz_pca_var(PCA, col.var = "black")
```

```{r}
# Visualize the squared cosines of variables in PCA
fviz_cos2(PCA, choice = "var", axes = 1:4)
```
```{r}
# Factor Analysis using factanal
numerical_nhanes_data.fa<-factanal(numerical_nhanes_data,factors = 4)
numerical_nhanes_data.fa
```
```{r}
# Compute squared loadings
apply(numerical_nhanes_data.fa$loadings^2,1,sum)
```
```{r}
# Factor Analysis using fa function
nhanes_data_PC<- fa(covariance_matrix ,nfactors = 4,rotate = "varimax",n.obs
= 1000 ,covar = TRUE,fm = "pa")
nhanes_data_PC
```

```{r}
# Rotate the factors
nhanes_data_PC_rotate <- fa(cor_matrix ,nfactors = 3,rotate =
"varimax",n.obs = 1000 ,cor = TRUE,fm = 'pa',max.iter = 1000)
nhanes_data_PC_rotate
```

```{r}
# Plot the factor diagram
fa.diagram(nhanes_data_PC_rotate)
```

```{r}
###Confirmatory Factor Analysis (CFA):
# Define the CFA model
variables <-
normalized_data[,c("RIDAGEYR","BMXBMI","LBXGLU","LBXGLT","SEQN")]
#define the CFA model
model <- '
Factor1 =~ SEQN
Factor2 =~ LBXGLT+LBXGLU+RIDAGEYR
Factor3=~BMXBMI
'
# Fit the CFA model
fit <- cfa(model, data = variables)
# Assess model fit
summary(fit, fit.measures = TRUE)
# Standardized estimates (factor loadings)
parameterEstimates(fit, standardized = TRUE, ci = TRUE)
```
```{r}
#remotes::install_github("rstudio/htmltools")

#install.packages("html",dependencies = TRUE)

```

